---
title: "artbin examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{artbin examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(artbin)
```

This vignette reproduces the examples from Section 3 of the artbin Stata Journal
article. Each section demonstrates a different study design.

## Binary outcome and comparison with published sample size

We reproduce the sample size calculation in Pocock (1983) for a 2-arm superiority
trial comparing anturan with placebo in patients after a myocardial infarction.
The primary outcome was death from any cause within one year. The control
(placebo) arm was expected to have a 10% probability of death and the treatment
arm a 5% probability, with the trial powered at 90%.

`artbin` infers the outcome direction from the probabilities: because the
treatment arm probability (0.05) is lower than the control (0.10), the outcome
is unfavourable (the trial aims to *reduce* the event probability).

The Wald test is used here because Pocock used the sample estimate of the null
variance.

```{r pocock}
artbin(pr = c(0.1, 0.05), alpha = 0.05, power = 0.9, wald = TRUE)
```

A total sample size of 1156 participants is required, matching the published
figure. The same result is obtained using `pr = c(0.9, 0.95)` with a favourable
outcome (survival).

## Binary outcome and comparison with the score test

`artbin` uses the score test by default. We compare it with the equivalent
call using the Wald test from the previous section.

```{r score}
artbin(pr = c(0.1, 0.05), alpha = 0.05, power = 0.9)
```

Both give a total sample size of 1164 with the score test.

## One-sided non-inferiority trial

A one-sided non-inferiority trial assuming a 90% probability of survival in both
arms, with the treatment arm required to be no more than 5 percentage points less
effective than control.

```{r ni-onesided}
artbin(pr = c(0.9, 0.9), margin = -0.05, onesided = TRUE)
```

A sample size of 457 is required in each group (914 total).

## Superiority trial with multiple arms

`artbin` supports trials with more than two arms. Rather than comparing each
treatment arm to control individually, it uses a global test for any difference
among the groups.

```{r multiarm}
artbin(pr = c(0.1, 0.2, 0.3, 0.4), alpha = 0.1, power = 0.9)
```

A sample size of 44 is required in each of the four groups (176 total).

## Complex non-inferiority trial: the STREAM trial

The STREAM trial evaluated a shorter intensive treatment for multi-drug resistant
tuberculosis against the lengthier WHO-recommended treatment. The design
parameters were:

- 70% favourable outcome probability on control, 75% on treatment
- 10 percentage-point non-inferiority margin ($m = \pi_2 - \pi_1 \geq -0.1$)
- 1:2 allocation (twice as many patients in the treatment arm)
- 20% loss to follow-up
- Wald test (commonly used in non-inferiority trials)

```{r stream}
artbin(
  pr      = c(0.7, 0.75),
  margin  = -0.1,
  power   = 0.8,
  aratios = c(1, 2),
  wald    = TRUE,
  ltfu    = 0.2
)
```

A total sample size of 398 is required (133 control, 265 treatment), allowing
for 20% loss to follow-up. When the STREAM trial concluded, it found that the
shorter treatment was only 1% less effective than the WHO-recommended treatment,
demonstrating significant evidence of non-inferiority.

## References

Pocock SJ (1983). *Clinical Trials: A Practical Approach*. Wiley.

Nunn AJ et al. (2019). A Trial of a Shorter Regimen for Rifampin-Resistant
Tuberculosis. *New England Journal of Medicine*, 380, 1201–1213.
